---
title: "How to run some compositional methods on data from QIIME2/DADA2"
author: "carina.bunse@hifmb.de and daniel.lundin@lnu.se"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    fig_caption: yes
    fig_height: 9
    fig_width: 8
    number_sections: yes
    toc: yes
  html_document:
    toc: yes
---

```{r setup, echo=F, cache = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path='figures/', cache = TRUE)
ggplot2::theme_set(ggplot2::theme_bw())
```

```{r libraries, message = FALSE, cache = FALSE}
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(breakaway))
```

```{r constants}
```

# Introduction

```{r read-data}
counts <- read_tsv(
  '../processed_data/feature-table.tsv', 
  skip = 1, 
  col_types = cols(.default = col_double(), `#OTU ID` = col_character())
) %>%
  rename(asv = `#OTU ID`) %>%
  gather(sample, count, 2:ncol(.)) %>%
  filter(count > 0) %>%
  group_by(sample) %>% filter(sum(count) > 100) %>% ungroup()
```

## Estimate actual richness

The `breakaway` package (https://rdrr.io/cran/breakaway/) supplies methods to esimate the true species richnes from
compositional count data, without rarefying. It gives you a point estimate plus an estimate of the standard error.

It's not so accessible for a biologist, however. Instead of starting with a table consisting of counts per ASV and
sample and giving you an estimated richness per sample, you have to: 1) do it once per sample and 2) start with a
table that consists of "counts of counts". With counts of counts I mean that the algorithm wants to know how many 
times an ASV with a certain number, 10 say, observations were found. Moreover, the first line needs to be the
number of times singletons were observed, i.e. how many times did you see a sequence that only appeared once in the
sample.

```{r estimate-richness}
# Create a table that will contain species richness for each sample
est_richness <- tibble(sample = character(), richness_est = double(), richness_se_est = double())

# Loop over all samples
for ( s in counts %>% distinct(sample) %>% pull(sample) ) {
  # Calculate estimated species richness for the sample `s`
  r <- counts %>% filter(sample == s) %>%
    # Create a "frequency count" table required by the Willis/Bunge methods
    group_by(count) %>%
    summarise(n = n()) %>%
    ungroup() %>%
    arrange(count) %>%
    # Turn it into a data.frame, because tibbles didn't work
    as.data.frame() %>%
    # Calculate the richness
    breakaway(print = FALSE, plot = FALSE, answers = TRUE)
  
  # Often, the algorithm doesn't converge, i.e. produce a result so we have to test if `r` is null before trying to add
  # it to the table.
  if ( ! is.null(r) ) {
    # Add the current result to the growing table
    est_richness <- est_richness %>%
      union(
        tibble(sample = s, richness_est = r$est, richness_se_est = r$seest)
      )
  }
}
  
  # Plot
  est_richness %>%
    mutate(min = richness_est - richness_se_est, max = richness_est + richness_se_est) %>%
    ggplot(aes(x = sample, y = richness_est)) +
    geom_point() +
    geom_errorbar(aes(ymin = min, ymax = max))
```

```{r add-clr}
# Add clr values
counts <- counts %>%
  right_join(
    counts %>% 
      spread(sample, count, fill = 0) %>% 
      data.frame() %>% tibble::column_to_rownames('asv') %>% 
      cmultRepl(method = 'CZM', output = 'p-counts') %>% 
      codaSeq.clr(samples.by.row = TRUE) %>%
      data.frame() %>%
      tibble::rownames_to_column('asv') %>% gather(sample, clr, 2:ncol(.)),
    by = c('asv', 'sample')
  ) %>%
  replace_na(list(count = 0))
```

```{r aitchinson-pca}
pca <- counts %>% 
  dplyr::select(asv, sample, clr) %>% 
  spread(asv, clr) %>% data.frame() %>% tibble::column_to_rownames('sample') %>% 
  dist(diag = TRUE) %>% 
  prcomp(scale. = TRUE)

propexpl <- pca$sdev^2/sum(pca$sdev^2)

broom::tidy(pca) %>% mutate(PC = sprintf("PC%d", PC)) %>%
  spread(PC, value) %>%
  ggplot(aes(x = PC1, y = PC2)) +
  geom_point() +
  xlab(sprintf("PC1 (%2.0f%%)", propexpl[1] * 100)) +
  ylab(sprintf("PC2 (%2.0f%%)", propexpl[2] * 100)) 
```

